\begin{frame}{Adaptation}

To transition from trees to flat sets, values must be moved from nodes directly into descriptors. It should consider three phases:

\begin{itemize}
    \item \textbf{Acquire:} Thread installs its descriptor $D$ into target cells $R_j$ via CAS in ascending address order (prevents deadlocks). If blocked by descriptor $U$, the thread helps $U$ finish first.

    \note{Modifying a frame now required marking all participating vertices via CAS, executing an atomic CAS replacement, and subsequently releasing them for subsequent operations}

    \pause

    \item \textbf{Decision (Linearization Point):} Once all cells are acquired, the thread atomically changes descriptor status via CAS from \textit{Undecided} to \textit{Committed} or \textit{Aborted}.

    \pause

    \item \textbf{Release:} Thread resets cell states to concrete values. Updates to $new_j$ if \textit{Committed}, or reverts to $old_j$ if \textit{Aborted}, exposing the changes to the system.
\end{itemize}

\pause

Since the specification draws no difference between threads and their descriptors, all updates occur concurrently.

\note{This means that it is built upon operations whose unique sequence strictly determines from the state of the descriptor}

\end{frame}


\begin{frame}{Algorithm Specifications}

State variables:

\begin{itemize}
    \item \texttt{wDescState} Descriptor status (\textit{IDLE, ACTIVE, SCANNING, SUCCESS, ABORTED}).
    \item \texttt{wDescPending} Cells awaiting locking or physical update.
    \item \texttt{wDescFrame} Keys (cells) acquired by the transaction.
    \item \texttt{wDescOldValue} Snapshot of original cell values ($old_j$).
    \item \texttt{wDescNewValue} Snapshot of target cell values ($new_j$).
\end{itemize}

\note{The specification uses the following temporal variables that change over time based on the \texttt{Next} predicate. This predicate is built from a set of conditions that refine the behavior of phases, whose described on previous slide}

\note{Executing operations requires helping other threads, which degrades read-only transaction performance. Therefore, readers simply announce their presence, and writers track these announcements}

\end{frame}
